```{r}
data('airquality')
airquality
airquality=airquality[complete.cases(airquality),]
summary(airquality)

plot(airquality$Ozone)
```

## Anova

```{r}
model <- lm(Ozone ~ Solar.R + Wind + Temp + Month + Day, data = airquality)

model
summary(model)

```

---
## Diagnostic
- Residuals vs Fitted. Used to check the linear relationship assumptions. A horizontal line, without distinct patterns is an indication for a linear relationship, what is good.

- Normal Q-Q. Used to examine whether the residuals are normally distributed. It’s good if residuals points follow the straight dashed line.

- Scale-Location (or Spread-Location). Used to check the homogeneity of variance of the residuals (homoscedasticity). Horizontal line with equally spread points is a good indication of homoscedasticity. 

- Residuals vs Leverage. Used to identify influential cases, that is extreme values that might influence the regression results when included or excluded from the analysis.

---

```{r, fig.width=8, fig.height=8}
par(mfrow = c(2, 2))
plot(model,)
```

---
### Cook's distance to determine the influence of a value.
This metric defines influence as a combination of leverage and residual size.
---

```{r}
plot(model, 4)
```


#homework:
1. What can you do to improve the model? 
2. Y = beta*X + e with X=[1,X1,X2,X3,...], what do you think if we
- add X1 in X so that X=[1,X1,X1,X2,X3,...]
- repeat each sample twice 












```{r}
a1=airquality
a1$Temp1=a1$Temp
model1 <- lm(Ozone ~ Solar.R + Wind + Temp + Temp1 + Month + Day, data = a1)
model1
summary(model1)
```


```{r}
a2=rbind(airquality,airquality)
a2
model2 <- lm(Ozone ~ Solar.R + Wind + Temp + Month + Day, data = a2)
model2
summary(model2)
```




