---
title: "R GLM - case studies 1) binary diabetes data diabetes 2) count data England covid-19"
output: html_notebook
---

```{r}
library(tidyverse)
library(sf) 
library(sp) 
library(lubridate)
library(FRK)
library(mlbench)

```

## Predict diabetes

```{r}
# load R built in data
data(PimaIndiansDiabetes)
PimaIndiansDiabetes
```
```{r}
logit_m = glm(diabetes ~ . + pressure*glucose, data = PimaIndiansDiabetes, family = binomial(link = "logit"))
logit_m %>% summary()

```

### The *null deviance* tells us how well the response variable can be predicted by a model with only an intercept term.

### The *residual deviance* tells us how well the response variable can be predicted by a model with p predictor variables. The lower the value, the better the model is able to predict the value of the response variable.




## Predict covid cases in England

```{r}
# read uk local authority boundary shapefile
utla_shp <- st_read("data/ons_utla.shp") 

# read census and IMD data
censusimd <- read_csv("data/2011census_2019imd_utla.csv")

# read uk covid data 2020-2021
covid19 <- read_csv("data/covid19_cases.csv")
head(covid19, 10)

```



```{r}
covid19 <- read_csv("data/covid19_cases.csv")

# join dfs
covid19_spt <- left_join(utla_shp, covid19, by = c("ctyu19nm" = "Area.name"))

# select pop data
pop <- censusimd %>% dplyr::select("UTLA19NM", "Residents", "Longterm_sick_or_disabled")
# join dfs
covid19_spt <- left_join(covid19_spt, pop,
                         by = c("ctyu19nm" = "UTLA19NM"))

# rate of new covid-19 infection
covid19_spt$n_covid19_r <- round( (covid19_spt$Daily.lab.confirmed.cases / covid19_spt$Residents) * 100000)

# risk of cumulative covid-19 infection
covid19_spt$c_covid19_r <- round( (covid19_spt$Cumulative.lab.confirmed.cases / covid19_spt$Residents) * 100000)

# share of population in long-term illness 
covid19_spt <- covid19_spt %>% mutate(
 lt_illness = Longterm_sick_or_disabled / Residents
)

# selecting variables
reg_df <- covid19_spt %>%
  dplyr::select(n_covid19_r, long, lat, date, lt_illness)

reg_df <- reg_df%>% st_drop_geometry()
reg_df

```

```{r}
eq1 <- n_covid19_r ~ .
lm_m <- lm(formula = eq1, data = reg_df)
lm_m %>% summary()
```

```{r}
# estimate a poisson model
poisson_m1 <- glm(eq1, family = poisson("log"), # Poisson + log link
                data = reg_df)
poisson_m1 %>% summary()
```

```{r}
# overdispersed poisson example
set.seed(999)
hist(Y1 <- rpois(1000, 2), breaks=seq(0, 30), col="orange", freq=F, ylim=c(0, 0.45), las=1, main="", xlab="Y")
hist(Y2 <- rpois(1000, 2 * exp( rnorm(1000, mean=0, sd=1))), add=T, freq=F, breaks=seq(0, 100))
legend("right", legend=c("Poisson", "overdispersed Poisson"), pch=15, col=c("orange", "grey50"), bty="n", cex=1.5)
```



## Poisson model assume mean = variance, but the results suggest an overdispersion.

```{r}
performance::check_overdispersion(poisson_m1)
```



```{r}
# estimate a quasipoisson model
qpoisson_m1 <- glm(eq1,
                family = quasipoisson("log"), # QuasiPoisson + log link
                data = reg_df)
qpoisson_m1 %>% summary()
```

## The quasi-Poisson assumes that the variance is proportional to the mean (the over-dispersion parameter). The model only changes the standard error of the estimated coefficients, but does not improve in terms of residual deviance.



```{r}
# estimate a negative binomial model

eq2 <- n_covid19_r ~ . + long*lat + lt_illness*long 

nb_m1 <- glm.nb(eq2, data = reg_df)
nb_m1 %>% summary()
```

# The NBM leads to a major improvement (reduction in residual deviance).


# Additional info: spatio decomposition
```{r}
# build basis functions
G <- auto_basis(data = reg_df[,c("long","lat")] %>%
                       SpatialPoints(),           
                nres = 1,                         
                type = "Gaussian")   

# basis functions evaluated at data locations are then the covariates
S <- eval_basis(basis = G,                       # basis functions
                s = reg_df[,c("long","lat")] %>%
                     as.matrix()) %>%            
     as.matrix()                                 
colnames(S) <- paste0("B", 1:ncol(S)) 

reg_df_st <- cbind(reg_df, S)
reg_df_st

```

# Homework: visualize spatial components B1-B9 and do GLM by including B1-B9




